---
title: "Introduction to `two2tango`"
vignette: >
  %\VignetteIndexEntry{Example Gaussian response}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(two2tango)
library(spatstat)
library(tmap)
tmap_mode("plot")
library(terra)
library(gstat)
library(sf)
library(tidyverse)
```

`two2tango` simulates pairs of virtual species with a known, adjustable degree
of spatial association, for testing and benchmarking species co-occurrence
and association methods against a known ground truth.

## How it works

1. An environmental predictor (a `SpatRaster` or spatstat `im`) defines the
   landscape.
2. Each species' intensity is modeled as a function of that predictor —
   Gaussian in `two2tango()`, linear in `two2waltz()`.
3. Correlated Gaussian noise (controlled by `var.e` and `cov.e`) is added to
   both species' log-intensity surfaces, inducing spatial association
   between them independent of their individual environmental responses.
4. Occurrence points for each species are sampled via an inhomogeneous
   Poisson process (`spatstat.random::rpoispp()`) weighted by the resulting
   intensity surface.
5. The true correlation between the species' log-intensity surfaces — the
   association driving the simulation — is returned alongside the simulated
   points, so it can be compared against whatever you estimate from the
   sampled data.

## Gaussian response

### Model for the virtual species

The model assumes an underlying Poisson point pattern, and a Gaussian response:

$\lambda_1 = peak_1 \times exp^{(-1/2 \times \dfrac{(predictor - \mu_1)^2}{\sigma_1^2} + e_1)}$  
$\lambda_2 = peak_2 \times exp^{(-1/2 \times \dfrac{(predictor - \mu_2)^2}{\sigma_2^2} + e_2)}$  

$e_{ij} \sim \textsf{MVN}(0, \Sigma)$  

Where, $\Sigma = \begin{bmatrix} var_{1,1} & cov_{1,2} \\ cov_{2,1} & var_{2,2} \end{bmatrix}$  

The inverse of the covariance matrix is called the precision matrix, denoted by $\tau = {\Sigma}^{-1}$.

### Function

The function `two2tango()` needs the following arguments:

  - `mu_1` = $\mu_1$ and `mu_2` = $\mu_2$: the mean of the response curve (niche mean) for each species,   
  - `sigma_1` = $\sigma_1$ and `sigma_2` = $\sigma_2$: the SD of the response curve (niche breath) for each species,  
  - `peak_1` = $peak_1$ and `peak_2` = $peak_2$: constants that set the expected abundance at the mean,   
  - `var` = $var_{1,1}$ = $var_{2,2}$: the variance of each species, which will always be set to `1`, 
  - `cov` = $cov_{1,2}$ and $cov_{2,1}$: the covariance of one species against the other, which needs to be symmetric.  

Then it returns a list with two `sf` objects of `POINT` geometry, one for each species, and the Pearson correlation at the level of the Poisson process.

:::{.callout-note}
The function `as.im.SpatRaster()` is not yet on CRAN for `spatstat`. See question in Stack overflow [Convert raster (terra) to im object (spatstat)](https://stackoverflow.com/questions/77916041/convert-raster-terra-to-im-object-spatstat).
:::

## Test the function

We will use as an example predictor the average annual temperature for Uruguay

```{r}
#| label: predictor
#| message: false
#| warning: false

uruguay <- geodata::gadm(country = 'UY', level=0, path = 'data/')
temperature <- geodata::worldclim_country('UY', var = 'tavg', path = 'data/')
temperature <- mean(temperature, na.rm=T) %>% mask(uruguay)
temp <- scale(temperature) %>% 
  aggregate(10)


tm_shape(temp) + 
  tm_raster(col.scale = tm_scale_continuous(midpoint = NA, values = 'brewer.rd_bu'), 
            col.legend = tm_legend('temperature')) +
tm_shape(uruguay) +
  tm_borders() + 
  tm_layout(frame=F, legend.frame = F)
```

### Case 1

Species have the same niche and co-occur

|       | $\mu_1$  | $\sigma_1$ | $peak_1$ | $var$ | $cov$ |
|-------|----------|------------|----------|-------|-------|
| `sp1` | 0.5      | 0.5        | 60       | 1     | 0.9   |
| `sp2` | 0.5      | 0.5        | 60       | 1     | 0.9   |


```{r}
#| label: gaussian-case1

mu1 = 1
mu2 = 1
sigma1 = 0.5
sigma2 = 0.5
peak1 = 0.01
peak2 = 0.01
cov.e= 0.9
var.e = 1

simulated_species <- two2tango(peak1=peak1, peak2=peak2,
                               mu1=mu1, sigma1=sigma1,
                               mu2=mu2, sigma2=sigma2,
                               cov.e=cov.e, var.e=var.e,
                               predictor = temp)

sp1 <- simulated_species[[1]] %>% mutate(species = 'sp1')
sp2 <- simulated_species[[2]] %>% mutate(species = 'sp2')

cor.pp <-  simulated_species[[3]]

tm_shape(temp) + 
  tm_raster(col.scale = tm_scale_continuous(midpoint = NA, values = 'brewer.rd_bu'), 
            col.legend = tm_legend('')) +
tm_shape(uruguay) + tm_borders() +
tm_shape(sp1) + 
  tm_dots(fill='species', 
          fill.scale = tm_scale_categorical(values='grey90'), 
          fill.legend = tm_legend('', frame=F), size = 0.5) +
tm_shape(sp2) + 
  tm_dots(fill='species', 
          fill.scale = tm_scale_categorical(values='black'),
          fill.legend = tm_legend(''), size = 0.5) + 
  tm_layout(frame=F, legend.frame = F) +
   tm_title(text = str_glue('cor = {round(cor.pp$estimate,3)}'))
```

The correlation at the level of the Poisson process between species is r = `r round(cor.pp$estimate,3)`.

```{r}
#| label: gaussian-case1-plot
#| code-fold: true

spec.response.tango <- function(x, mu, peak, sigma) {
    # peak1*exp(-0.5*((predictor-mu1)^2)/(sigma1^2) + e[,1])
    y <- peak*exp(-0.5*((x-mu)^2)/(sigma^2))
    return(y)
}

response.df <- tibble(x = seq(-3, 3, by = 0.01),
                      y1 = spec.response.tango(x, mu1, peak1, sigma1),
                      y2 = spec.response.tango(x, mu2, peak2, sigma2))

ggplot() +
    geom_line(data=response.df, aes(x=x, y=y1), col='red', linetype = 'dashed') + 
    geom_point(data=response.df, aes(x=x, y=y1), col='red') + 
    geom_line(data=response.df, aes(x=x, y=y2), col='black') + 
    geom_line(data=response.df, aes(x=x, y=y2), col='black') + 
    labs(y='Y') + theme_bw()
```



### Case 2

Species have a different niche and negative co-occurrence

|       | $\mu_1$  | $\sigma_1$ | $peak_1$ | $var$ | $cov$ |
|-------|----------|------------|----------|-------|-------|
| `sp1` | 0.25     | 0.5        | 60       | 1     | -0.9  |
| `sp2` | -0.25    | 0.5        | 60       | 1     | -0.9  |


```{r}
#| label: gaussian-case2

mu1 = 1
mu2 = -1
sigma1 = 0.5
sigma2 = 0.5
peak1 = 0.01
peak2 = 0.01
cov.e = -0.9
var.e = 1

simulated_species <- two2tango(peak1=peak1, peak2=peak2,
                               mu1=mu1, sigma1=sigma1,
                               mu2=mu2, sigma2=sigma2,
                               cov.e=cov.e, var.e=var.e,
                               predictor = temp)

sp1 <- simulated_species[[1]] %>% mutate(species = 'sp1')
sp2 <- simulated_species[[2]] %>% mutate(species = 'sp2')

cor.pp <-  simulated_species[[3]]

tm_shape(temp) + 
  tm_raster(col.scale = tm_scale_continuous(midpoint = NA, values = 'brewer.rd_bu'), 
            col.legend = tm_legend('')) +
tm_shape(uruguay) + tm_borders() +
tm_shape(sp1) + 
  tm_dots(fill='species', 
          fill.scale = tm_scale_categorical(values='grey90'), 
          fill.legend = tm_legend('', frame=F), size = 0.5) +
tm_shape(sp2) + 
  tm_dots(fill='species', 
          fill.scale = tm_scale_categorical(values='black'),
          fill.legend = tm_legend(''), size = 0.5) + 
  tm_layout(frame=F, legend.frame = F) +
   tm_title(text = str_glue('cor = {round(cor.pp$estimate,3)}'))
```


```{r}
#| label: gaussian-case2-plot
#| code-fold: true

response.df <- tibble(x = seq(-3, 3, by = 0.01),
                      y1 = spec.response.tango(x, mu1, peak1, sigma1),
                      y2 = spec.response.tango(x, mu2, peak2, sigma2))

ggplot() +
    geom_line(data=response.df, aes(x=x, y=y1), col='red', linetype = 'dashed') + 
    geom_point(data=response.df, aes(x=x, y=y1), col='red') + 
    geom_line(data=response.df, aes(x=x, y=y2), col='black') + 
    geom_line(data=response.df, aes(x=x, y=y2), col='black') + 
    labs(y='Y') + theme_bw()
```


## Linear response

### Model for the virtual species

The model assumes an underlying Poisson point pattern, and a Gaussian response:

$\lambda_1 = exp^{\alpha_1 + \beta_1 \times predictor + e_1)}$   
$\lambda_2 = exp^{\alpha_2 + \beta_2 \times predictor + e_2)}$   

$e_{ij} \sim \textsf{MVN}(0, \Sigma)$  

Where, $\Sigma = \begin{bmatrix} var_{1,1} & cov_{1,2} \\ cov_{2,1} & var_{2,2} \end{bmatrix}$  

The inverse of the covariance matrix is called the precision matrix, denoted by $\tau = {\Sigma}^{-1}$.  

### Function

The function `two2waltz()` needs the following arguments:

  - `alpha_1` = $\alpha_1$ and `alpha_2` = $\alpha_2$: the intercept of the response (niche mean) for each species,   
  - `beta_1` = $\beta_1$ and `beta_2` = $\beta_2$: the SD of the response curve (niche breath) for each species,  
  - `var` = $var_{1,1}$ = $var_{2,2}$: the variance of each species, which will always be set to `1`, 
  - `cov` = $cov_{1,2}$ and $cov_{2,1}$: the covariance of one species against the other, which needs to be symmetric.  

Then it returns a list with two `sf` objects of `POINT` geometry, one for each species, and the Pearson correlation at the level of the Poisson process.

## Test the function

We will use the same predictor as before, the average annual temperature for Uruguay.  

```{r}
#| label: predictor-plot
#| message: false
#| warning: false

tm_shape(temp) + 
  tm_raster(col.scale = tm_scale_continuous(midpoint = NA, values = 'brewer.rd_bu'), 
            col.legend = tm_legend('temperature')) +
tm_shape(uruguay) +
  tm_borders() + 
  tm_layout(frame=F, legend.frame = F)
```

### Case 1

Species have the same niche and co-occur

|       | $\alpha_1$ | $\beta_1$ | $var$ | $cov$ |
|-------|------------|-----------|-------|-------|
| `sp1` | -1         | 0.01      | 1     | 0.9   |
| `sp2` | -1         | 0.01      | 1     | 0.9   |


```{r}
#| label: linear-case1

alpha1 = -2
alpha2 = -2
beta1 = 1
beta2 = 1
cov.e = 0.9
var.e = 1

simulated_species <- two2waltz(alpha1=alpha1, alpha2=alpha2,
                            beta1=beta1, beta2=beta2,
                            cov.e=cov.e, var.e=var.e,
                            predictor = temp)

sp1 <- simulated_species[[1]] %>% mutate(species = 'sp1')
sp2 <- simulated_species[[2]] %>% mutate(species = 'sp2')

cor.pp <-  simulated_species[[3]]

tm_shape(temp) + 
  tm_raster(col.scale = tm_scale_continuous(midpoint = NA, values = 'brewer.rd_bu'), 
            col.legend = tm_legend('')) +
tm_shape(uruguay) + tm_borders() +
tm_shape(sp1) + 
  tm_dots(fill='species', 
          fill.scale = tm_scale_categorical(values='grey90'), 
          fill.legend = tm_legend('', frame=F), size = 0.5) +
tm_shape(sp2) + 
  tm_dots(fill='species', 
          fill.scale = tm_scale_categorical(values='black'),
          fill.legend = tm_legend(''), size = 0.5) + 
  tm_layout(frame=F, legend.frame = F) +
   tm_title(text = str_glue('cor = {round(cor.pp$estimate,3)}'))
```

The correlation at the level of the Poisson process between species is r = `r round(cor.pp$estimate,3)`.

```{r}
#| label: linear-case1-plot
#| code-fold: true

spec.response.waltz <- function(x, alpha, beta) {
  # exp(alpha1 + beta1*predictor + e[,1])
  y <- exp(alpha + beta*x)
  return(y)
}

response.df <- tibble(x = seq(-3, 3, by = 0.01),
                      y1 = spec.response.waltz(x, alpha1, beta1),
                      y2 = spec.response.waltz(x, alpha2, beta2))

ggplot() +
    geom_line(data=response.df, aes(x=x, y=y1), col='red', linetype = 'dashed') + 
    geom_point(data=response.df, aes(x=x, y=y1), col='red') + 
    geom_line(data=response.df, aes(x=x, y=y2), col='black') + 
    geom_line(data=response.df, aes(x=x, y=y2), col='black') + 
    labs(y='Y') + theme_bw()
```

### Case 2

Species have a different niche and negative co-occurrence

|       | $\alpha_1$ | $\beta_1$ | $var$ | $cov$ |
|-------|------------|-----------|-------|-------|
| `sp1` | -2         | 1         | 1     | -0.9  |
| `sp2` | -2         | -1        | 1     | -0.9  |


```{r}
#| label: linear-case2

alpha1 = -2
alpha2 = -2
beta1 = 1
beta2 = -1
cov.e = -0.9
var.e = 1

simulated_species <- two2waltz(alpha1=alpha1, alpha2=alpha2,
                            beta1=beta1, beta2=beta2,
                            cov.e=cov.e, var.e=var.e,
                            predictor = temp)

sp1 <- simulated_species[[1]] %>% mutate(species = 'sp1')
sp2 <- simulated_species[[2]] %>% mutate(species = 'sp2')

cor.pp <-  simulated_species[[3]]

tm_shape(temp) + 
  tm_raster(col.scale = tm_scale_continuous(midpoint = NA, values = 'brewer.rd_bu'), 
            col.legend = tm_legend('')) +
tm_shape(uruguay) + tm_borders() +
tm_shape(sp1) + 
  tm_dots(fill='species', 
          fill.scale = tm_scale_categorical(values='grey90'), 
          fill.legend = tm_legend('', frame=F), size = 0.5) +
tm_shape(sp2) + 
  tm_dots(fill='species', 
          fill.scale = tm_scale_categorical(values='black'),
          fill.legend = tm_legend(''), size = 0.5) + 
  tm_layout(frame=F, legend.frame = F) +
   tm_title(text = str_glue('cor = {round(cor.pp$estimate,3)}'))
```

The correlation at the level of the Poisson process between species is r = `r round(cor.pp$estimate,3)`.

```{r}
#| label: linear-case2-plot
#| code-fold: true

response.df <- tibble(x = seq(-3, 3, by = 0.01),
                      y1 = spec.response.waltz(x, alpha1, beta1),
                      y2 = spec.response.waltz(x, alpha2, beta2))

ggplot() +
    geom_line(data=response.df, aes(x=x, y=y1), col='red', linetype = 'dashed') + 
    geom_point(data=response.df, aes(x=x, y=y1), col='red') + 
    geom_line(data=response.df, aes(x=x, y=y2), col='black') + 
    geom_line(data=response.df, aes(x=x, y=y2), col='black') + 
    labs(y='Y') + theme_bw()
```
